\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-22 tex/paper.tex
Fixed Empirical Bayes confusion

\subsection{Empirical Bayes through  Variational Inference}

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the parametrized marginal distribution of the generated data
\begin{equation}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
in \citep{2016arXiv161009033R}, though more complicated than a simple
reparametrization of normally distributed noise $\bfz$ through a differentiable
function. In our case, the variational program is a
marginalized, non-differentiable  simulator.  Its density is  intractable;
nevertheless, it can generate samples for $\bfx$ whose expectations are differentiable with
respect to $\bfpsi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

%Typically in variational inference one would has access to the posterior $p(\bftheta | \bfx )$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.

Typically in variational inference one starts with a tractable prior
$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
approximate the posterior, one would criticize the model through the
posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
In our case, we can identify the proposal distribution with the variational posterior
$q(\bftheta | \bfpsi)$, which we effectively optimize by
directly minimizing the  divergence of the corresponding posterior predictive
distribution $\qxpsi$ and the data distribution $p_r(\bfx)$, without evaluating
the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
Consequently, the proposed algorithm is a form of variational inference through
model criticism~\citep{box1980sampling}.

We can also view the optimization  through the lens of empirical Bayes (EB),
where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
%posterior predictive distribution.
Traditionally, we would have access to the likelihood and optimize with respect to the family of posteriors
 $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi) d\bftheta $
as an intermediate step to computing $\bfpsi^*$. In our case, the likelihood is intractable and we have not explicitly
specified a prior family. However, the procedure corresponds to empirical Bayes if we identify
$q(\bftheta | \bfpsi^*) = p(\bftheta | \bfx , \bfpsi)$. Thus we call our method AVO-EB and note that it is applicable even outside of the likelihood-free context.

Algorithm~\ref{alg:avo}, can be easily modified to have different sizes of mini-batches,
$M_r$ for the observed data and $M_g$ for the simulator. When $M_g \gg M_r$,
then the variance of $\bftheta^*$ scales as
$\sqrt{M_r}$~\citep{bernton2017inference}. A final $\sqrt{N/M}$ scaling to the
log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of
the data.

@@

\subsection{Point estimates}

%\kcnote{connect to MWE and MEWE here?}

When $\rho$ is the Wasserstein distance, $\bftheta^*$ is referred to as the
minimum Wasserstein estimator (MWE). When the model is well specified, $\bftheta^*$
coincides with the true data-generating parameter; however, if the model is
misspecified, $\bftheta^*$ is typically different from the maximum likelihood
estimator (MLE), which is the minimizer of $\text{KL}(p_r(\bfx) \parallel p(\bfx |
\bftheta))$~\citep{bernton2017inference}. In order to more effectively target
point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
regularization term corresponding to the differential entropy $H$ of the
proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
We refer to the algorithm with $\gamma>0$ as AVO-MWE.

For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the posterior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
    \caption{Electron--positron annihilation.
    ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization. The density of the penalized distribution ($\gamma=5$) is here highly concentrated, resulting in the green mass near $\bftheta^*$.
    ({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their respective proposal distributions, both models closely match the observed data.
    ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:weinberg}
\end{figure}

@@

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
AVO. We see that the distributions both arrive
in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\bftheta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
results for the entropy penalized model. This suggests either a relatively flat
landscape around $\bftheta^*$ or that the observed data can in this case also be
reproduced with the posterior predictive distribution $\qxpsi$.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution.
%In summary, despite the intricate
%relation between the observations $\mathbf{x}$, the latent variable $\bfz$
%and the parameters $\bftheta$, this example demonstrates the efficacy
%of AVO for parameter inference on a black box
%simulator.

>>>>

\subsection{Point Estimates through Marginal Likelihood}

%\subsection{Empirical Bayes through  Variational Inference}

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the parametrized marginal distribution of the generated data
\begin{equation}\label{eq:marginal_likelihood}
    \qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
in \citep{2016arXiv161009033R}, though more complicated than a simple
reparametrization of normally distributed noise $\bfz$ through a differentiable
function. In our case, the variational program is a
marginalized, non-differentiable  simulator.  Its density is  intractable;
nevertheless, it can generate samples for $\bfx$ whose expectations are differentiable with
respect to $\bfpsi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim \qxpsi \equiv \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

%where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
%%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
%%posterior predictive distribution.

%%Typically in variational inference one would has access to the posterior $p(\bftheta | \bfx )$ (or the joint density $p(\bfx,\bftheta)$), and after optimizing $q(\bftheta | \bfpsi)$ one would criticize the model through the the posterior predictive distribution as advocated by Box~\citep{box1980sampling}.
%
%Typically in variational inference one starts with a tractable prior
%$p(\bftheta)$ and a tractable likelihood $p(\bfx |\bftheta )$, thus fixing the
%posterior $p(\bftheta | \bfx )$.   After optimizing $q(\bftheta | \bfpsi)$ to
%approximate the posterior, one would criticize the model through the
%posterior predictive distribution as advocated by Rubin~\citep{rubin1984, box1980sampling}.
%In our case, we can identify the proposal distribution with the variational posterior
%$q(\bftheta | \bfpsi)$, which we effectively optimize by
%directly minimizing the  divergence of the corresponding posterior predictive
%distribution $\qxpsi$ and the data distribution $p_r(\bfx)$, without evaluating
%the intractable likelihood $p(\bfx | \bftheta )$ or specifying a prior.
%Consequently, the proposed algorithm is a form of variational inference through
%model criticism~\citep{box1980sampling}.
%
%%% KC: Confusion here about $q(\bftheta | \bfpsi)$ as posterior or prior.
%%% need differentiate xnew vs. xold and q_new and q_old
%%% Seems like there is still a route to get at likelihood or posterior
%%% Could be that issue is we loose normalizing constant between q_old(theta|psi) and likleihood
%
%We can also view the optimization  through the lens of empirical Bayes (EB),
%where one specifies a family of priors $p(\bftheta | \bfpsi)$ and then estimates $\bfpsi^*$ using the data.
%%In our case, we find the $\bfpsi^*$ that minimizes the Wasserstein distance between the data distribution and the
%%posterior predictive distribution.
%Traditionally, we would have access to the likelihood and optimize with respect to the family of posteriors
% $p(\bftheta | \bfx, \bfpsi) = p(\bfx | \bftheta) p(\bftheta | \bfpsi) / \int p(\bfx | \bftheta) p(\bftheta | \bfpsi) d\bftheta $
%as an intermediate step to computing $\bfpsi^*$. In our case, the likelihood is intractable and we have not explicitly
%specified a prior family. However, the procedure corresponds to empirical Bayes if we identify
%$q(\bftheta | \bfpsi) = p(\bftheta | \bfx , \bfpsi)$. Thus we call our method AVO-EB and note that it is applicable even outside of the likelihood-free context.

%%KC: Not necessary if we aren't interpreting as posterior
%Algorithm~\ref{alg:avo}, can be easily modified to have different sizes of mini-batches,
%$M_r$ for the observed data and $M_g$ for the simulator. When $M_g \gg M_r$,
%then the variance of $\bftheta^*$ scales as
%$\sqrt{M_r}$~\citep{bernton2017inference}. A final $\sqrt{N/M}$ scaling to the
%log-posterior $q(\bftheta | \bfpsi)$ is needed to reflect the full strength of
%the data.

@@

We can view the optimization of the marginal likelihood with respect to $\bfpsi$ through the lens of empirical Bayes (EB), where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$. If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\citep{rubin1984}.
When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
minimum Wasserstein estimator (MWE). 
When the model is well specified, the MWE
coincides with the true data-generating parameter; however, if the model is
misspecified, the MWE is typically different from the maximum likelihood
estimator (MLE).
Moreover, if the simulator $p(\bfx |
\bftheta)$ is misspecified, $q(\bftheta | \bfpsi)$ will attempt to smear it so that the marginal
model $q(\bfx | \bfpsi)$ is closer to $p_r(\bfx)$. However, if the simulator is well specified, 
then $q(\bftheta | \bfpsi)$ will concentrate its mass around the true data-generating parameter.

%When the model is well specified, $\bfpsi^*$
%coincides with the true data-generating parameter; however, if the model is
%misspecified, $\bftheta^*$ is typically different from the maximum likelihood
%estimator (MLE), which is the minimizer of $\text{KL}(p_r(\bfx) \parallel p(\bfx |
%\bftheta))$~\citep{bernton2017inference}.

In order to more effectively target
point estimates $\bftheta^*$, we augment Eqn.~\ref{eqn:vo-ug} with a
regularization term corresponding to the differential entropy $H$ of the
proposal distribution $q(\bftheta | \bfpsi)$. That is,
\begin{equation}
    U_g = \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [ {\cal L}_g ] + \gamma H(q(\bftheta|\bfpsi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
%We refer to the algorithm with $\gamma>0$ as AVO-MWE.

For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which results
in learning a smeared variation of the original simulator corresponding to the marginal or prior predictive model. On the other hand,
for large values of $\gamma$, the procedure is encouraged to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\bftheta$ values.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/weinberg.pdf}
    \caption{Electron--positron annihilation.
    ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after adversarial variational optimization.
                     Contours correspond to parameters $\bftheta$ within $1$-$2$-$3$ Mahalanobis distance units from the mean of $q(\bftheta|\bfpsi)$.
                     The density of the penalized distribution ($\gamma=5$) is  highly concentrated, resulting in the green mass near $\bftheta^*$.
    ({\it Top right}) Model distributions $\qxpsi$ after training. Despite the differences between their proposal distributions, both models closely match the observed data.
    ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:weinberg}
\end{figure}

@@

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
AVO. We see that the distributions both arrive
in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\bftheta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
results for the entropy penalized model. This suggests either a relatively flat
landscape around $\bftheta^*$ or that the observed data can in this case also be
reproduced with the  predictive distribution $\qxpsi$.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution.
%In summary, despite the intricate
%relation between the observations $\mathbf{x}$, the latent variable $\bfz$
%and the parameters $\bftheta$, this example demonstrates the efficacy
%of AVO for parameter inference on a black box
%simulator.
